A community-maintained standard library of population genetic models

### 2020-05-01 popSimManu.tex
Wordsmithing/polishing of Results
More minor changes, just polishing

Given the genome data and simulation model descriptions defined within the
library, it is then straightforward to run standardized simulations
across a range of organisms. \Stdpopsim has a Python API and a user-friendly
command line interface, allowing users with minimal experience direct access to
state-of-the-art simulations. Simulations are output in the ``tree sequence''
format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
contains complete genealogical information about the simulated samples, is
extremely compact, and can be processed efficiently using the \tskit library
\citep{kelleher2016efficient,kelleher2018efficient}.
The tree sequence format could also be converted
to other formats (e.g., VCF) by the user if desired.
Currently, \stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
as the default simulation engine.

Coalescent simulations, while highly efficient, are limited in their ability to model
complex selection scenarios, such as recurrent sweeps and background selection. For these reasons
we have also implemented the forward-time simulator, \texttt{SLiM} \citep{haller2019tree,haller2019slim},
as an alternative backend engine to \texttt{stdpopsim}, allowing for the simulation of processes that cannot be modeled under
the coalescent. However, as forward-time simulators explicitly model all individuals in a population, simulating large
population sizes can be highly demanding of computational resources. One common practice used to address this challenge is to
scale down the population size used in the simulations (\textit{N}),
scale down times specified in the demographic model by the same factor, and scale up the mutation and recombination
rates and selection coefficients such that the scaled parameters, \textit{N$\mu$}, \textit{Nr},
and \textit{Ns}, remain unchanged \citep[see the \texttt{SLiM} manual and][]{uricchio2014robust}.
Such down-scaled simulations are not completely equivalent to simulating
all individuals in the population, and may lead to subtle differences.
However, since many sequence-based measures of population diversity remain unchanged when rescaling in this fashion,
this practice is very common when using forward-time simulators, such as \texttt{SLiM}.

We validated our implementation of the \texttt{SLiM} engine by comparing
estimates of several population genetic summary statistics for neutral simulations generated by both \texttt{SLiM}
and \texttt{msprime}.
Examples of this validation for different down-scaling factors are shown in
Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}, which indicate that scaling down the simulated
population size by up to a factor
of $50$ ($Q=50$ in \texttt{SLiM}) produces patterns of both diversity and linkage disequilibrium that are
qualitatively similar to those observed under the coalescent (i.e. \texttt{msprime}), when both are simulated under a recombination map
(Figure \ref{fig:slim_val_map}).
We do observe a slight departure in values of Tajima's $D$ when scaling down by a factor of $50$,
likely due to the subtle differences produced by down-scaling.
Interestingly, this effect of down-scaling is more prominent and extends to other sequence-based statistcs
when data sets are simulated under a uniform recombination rate (Figure \ref{fig:slim_val_nomap}).
Importantly, both comparisons validate the equivalence of \texttt{SLiM} and \texttt{msprime} when no down-scaling is applied,
and they suggest that down-scaling population sizes can significantly reduce running time at a fairly modest cost to accuracy.
However, since the effects of rescaling are not fully understood, scaling factors should be used as conservatively as practicable.
The framework provided by \texttt{stdpopsim} should enable a deeper inspection of this issue, which we expect will contribute to
best practices used when applying forward-time simulators such as \texttt{SLiM}.

@@

The \stdpopsim command line interface, by default, outputs citation information
for the models, genetic maps, and simulation engines used in any particular run.
We hope that this will encourage users to appropriately acknowledge the
resources used in published work, and encourage authors
publishing demographic models to contribute to our ongoing community-driven development process.
Together with the \stdpopsim version number and the long-term stable identifiers
for population models and genetic maps,
this citation information will result in well-documented and reproducible
simulation workflows. The individual tree sequence files produced by
\stdpopsim also contain complete provenance information including the command
line arguments, operating system environment and versions of key libraries
used.

@@

As an example of the utility of \stdpopsim, we demonstrate how it can be used
to easily and fairly compare popular demographic inference methods.
Although we present comparison of results from several
methods, our aim at this stage is not to provide an exhaustive
evaluation or ranking of these methods. Our hope is instead to demonstrate how \stdpopsim
will facilitate more detailed future explorations of the strengths and weaknesses of the numerous
inference methods that are available to the population genetics community
(see Discussion).

We start by comparing popular methods for estimating
population size histories of single populations and subsequently
show simple examples of multi-population inference.
To reproducibly evaluate and compare the performance of inference methods, we developed
workflows using \texttt{snakemake} \citep{koster2012snakemake},
available from \url{https://github.com/popsim-consortium/analysis},
that allow efficient computing in multicore or cluster environments.
Our workflow generates $R$ replicates of $C$ chromosomes,
producing $n$ samples in each of a total of $R \times C$ simulations for each demographic model.
After simulation,
the workflow prepares input files for each inference method
by grouping all $n \times R \times C$ simulated chromosomes
into a single file.
Each file is then converted into an input file appropriate for each inference method
(such that all inference methods run on the same simulation replicates).
Each of the inference programs are then run in parallel, followed by plotting of
estimates of population size history from each program.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/HomSap_OutOfAfricaArchaicAdmixture_5R19.pdf}
\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
size over time ($N(t)$) inferred using 4 different methods: \smcpp, \texttt{stairway plot}, and
\MSMC with $n=2$ and $n=8$ samples. Data were generated by simulating
replicate human genomes under the OutOfAfricaArchaicAdmixture\_5R19 model \citep{ragsdale2019models} and using the
HapMapII\_GRCh37 genetic map~\citep{international2007second}. From top to bottom we show estimates for each
of the three populations in the model (YRI, CEU, and CHB). In shades of blue we show the estimated
$N(t)$ trajectories for each replicate.
As a proxy for the ``truth'', in black we show inverse coalescence rates
as calculated from the demographic model used for simulation (see text).
}
\label{fig:n_t_ragsdale}
\end{center}
\end{figure}

\subsubsection*{ Single-population demographic models.}
For single-population demographic models, we compared
\MSMC~\citep{schiffels2014inferring}, \smcpp~\citep{terhorst2017robust}, and
\stairwayplot~\citep{liu2015exploring}
 on simulated genomes sampled from a single population,
in a number of the demographic models described above.
When fitting single-population models to data simulated for multiple
populations with migration, we do not necessarily expect the inference methods to recover the
actual simulated population size due to model misspecification.
Indeed, the best one may expect in such a scenario is to infer a model which accurately reflects the
coalescence time distribution of the simulated model. While in the inferred single-population model,
this distribution is only affected by the ancestral effective population sizes, in the simulated multi-population
model it is also influenced by migration between the target population and populations not analyzed in inference.
Since ancestral population sizes correspond to the inverse of the coalescence rates, by computing the
inverse coalescence rates from the simulated model, we obtain the best estimates
one may hope to infer by analyzing a single population with a method that seeks to match coalescence time distributions.
. We thus analytically compute inverse coalescence
rates in \texttt{msprime} for each simulated model, and use them as a benchmark of the ``true'' effective population sizes.
See the Appendix for a precise definition and description of the inverse coalescence rate computation.

Figure \ref{fig:n_t_ragsdale} presents the results
from simulations under OutOfAfricaArchaicAdmixture\_5R19,
a model of human migration out of Africa that includes archaic admixture
\citep{ragsdale2019models}, along with an empirical genetic map. In each column of this figure
we show the inferred population size history (denoted $N(t)$)
from samples taken from each of the three extant populations in the model.
In each row we show comparisons among the methods (including two sample sizes for \MSMC).
Blue lines show estimates from each of three replicate whole genome simulations,
and black lines indicate the ``true" values depicted by the inverse coalescence rates (although in this specific model
the inverse coalescence rates are very close to the simulated population sizes;
Figure \ref{fig:census_vs_IRC}).
%DRS: trying to cut down on the redundancy between the coal rate stuff here and the previous paragraph
While there is variation in accuracy among methods, populations, and individual replicates,
the methods generally infer back a good estimate
of the true effective population sizes of the simulations --
inferred values are mostly within a factor of two of the truth,
and many of the methods infer a bottleneck at the correct time.

@@

We can also
see how well methods might do at recovering the population history of a constant-sized population,
with human genome architecture and genetic map.
We show results of such an
experiment in Figure \ref{fig:n_t_HomSap_constant}.
All methods recover population size within a factor of two of the simulated values, however
SMC-based methods, perhaps due to their regularization, tend to infer sinusoidal
patterns of population size even though no such change is present.

@@

\subsubsection*{Multi-population demographic models.}
As \stdpopsim implements multi-population demographic models, we also
explored parameter estimation of population divergence parameters. In particular,
we simulated data under multi-population models for humans and \textit{D.~melanogaster}
and then inferred  parameters using \dadi, \fastsimcoal, and \smcpp.
For simplicity, we conducted inference in \dadi and \fastsimcoal by fitting an isolation with migration (IM) model
with constant population sizes and bi-directional migration \citep{hey2004im}.
Our motivation for fitting this simple
IM model was to mimic the typical approach of two population inference on empirical
data, where the user is not aware of the `true' underlying demography and the inference
model is often misspecified.
For human models with more than two populations \citep[e.g.,][]{gutenkunst2009inferring}
this means that we are inferring parameters for a model that does
not match the model from which the data were generated (Figures
\ref{fig:IM_popn_human}A and B). However, since the model we use for inference also
allows gene flow between populations, we directly compare estimated effective population sizes
to the values used in simulations (black line in Figure \ref{fig:IM_popn_human}C)
and not the inverse coalescence rates.

In Figure \ref{fig:IM_popn_human}C we show estimates of population sizes and divergence
time, for each of the inference methods, using samples drawn from African and European populations
simulated under the OutOfAfrica\_3G09 model. Our results highlight many
of the strengths and weaknesses of the different types of methods.
For instance, the SFS-based approaches where we fit simple IM models do not capture
recent exponential growth in the CEU population, but do consistently recover the
simulated YRI population size history. Moreover, these approaches allow for estimating
migration rates (Figure \ref{fig:homsap_mig_rates}), also leading to more accurate inference
of divergence times. However, these migration rate estimates are somewhat biased,
likely due to model misspecification.
In contrast, \smcpp is much better at capturing the recent exponential
growth in the CEU population, though the inferred population sizes are generally noisier.
In addition, \smcpp consistently underestimates divergence times
because it assumes no migration between populations (Figure \ref{fig:IM_popn_human}C).

Again, we can compare between species and look at the performance of these methods
for a two-population model of \textit{D.~melanogaster}. Figure
\ref{fig:two_popn_fly} shows inference results using data simulated under
the OutOfAfrica\_2L06 model. This model includes an ancestral population in
Africa from which a European population splits off following a bottleneck, with no
post-divergence gene flow between the African and European population (Figure \ref{fig:two_popn_fly}A).
Here again, we find that \dadi and \fastsimcoal infer more consistent histories,
but they do not detect the brief bottleneck in Europe, due to inference model
not allowing for population size changes after the population split. In addition, \dadi
and \fastsimcoal both do reasonably well at correctly inferring the absence of
migration (Figure \ref{fig:dromel_mig_rates}). In contrast, the inferred
demographic parameters from \smcpp are more noisy, though in some cases better
capture the short bottleneck in the European population.

>>>>

Given the genome data and simulation model descriptions defined within the
library, it is straightforward to run standardized simulations
across a range of organisms. \Stdpopsim has a Python API and a user-friendly
command line interface, allowing users with minimal experience direct access to
state-of-the-art simulations. Simulations are output in the ``tree sequence''
format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
contains complete genealogical information about the simulated samples, is
extremely compact, and can be processed efficiently using the \tskit library
\citep{kelleher2016efficient,kelleher2018efficient}.
The tree sequence format could also be converted
to other formats (e.g., VCF) by the user if desired.
Currently, \stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
as the default simulation engine.

Coalescent simulations, while highly efficient, are limited in their ability to model
complex selection scenarios, such as recurrent sweeps and background selection. For these reasons,
we have also implemented the forward-time simulator, \texttt{SLiM} \citep{haller2019tree,haller2019slim},
as an alternative backend engine to \texttt{stdpopsim}, allowing for the simulation of processes that cannot be modeled under
the coalescent. However, as forward-time simulators explicitly model all individuals in a population, simulating large
population sizes can be highly demanding of computational resources. One common practice used to address this challenge is to
scale down the population size used in the simulations (\textit{N}),
scale down times specified in the demographic model by the same factor, and scale up the mutation and recombination
rates and selection coefficients such that the scaled parameters, \textit{N$\mu$}, \textit{Nr},
and \textit{Ns}, remain unchanged \citep[see the \texttt{SLiM} manual and][]{uricchio2014robust}.
Such down-scaled simulations are not completely equivalent to simulating
all individuals in the population, and may lead to subtle differences.
However, since many sequence-based measures of population diversity remain unchanged when rescaling in this fashion,
this practice is effective for many purposes and widely employed.

We validated our implementation of the \texttt{SLiM} engine by comparing
estimates of several population genetic summary statistics for neutral simulations generated by both \texttt{SLiM}
and \texttt{msprime}.
Examples of this validation for different down-scaling factors are shown in
Figures \ref{fig:slim_val_map} and \ref{fig:slim_val_nomap}, which indicate that scale factors of up to
$50$ ($Q=50$ in \texttt{SLiM}) produce patterns of both diversity and linkage disequilibrium that are
qualitatively similar to those observed under the coalescent (i.e. \texttt{msprime}), when both are simulated under a recombination map
(Figure \ref{fig:slim_val_map}).
We do observe a slight departure in values of Tajima's $D$ when scaling down by a factor of $50$.
%likely due to the subtle differences produced by down-scaling.
Interestingly, the effects of down-scaling are more prominent and extend to other sequence-based statistics
when data sets are simulated under a uniform recombination rate (Figure \ref{fig:slim_val_nomap}).
Importantly, both comparisons validate the equivalence of \texttt{SLiM} and \texttt{msprime} when no down-scaling is applied,
and they suggest that down-scaling population sizes can significantly reduce running time at a fairly modest cost to accuracy.
% ACS: We might consider deleting the next two sentences.  This feels like a lot on down-scaling, and it's a bit verbose as written.  Maybe 
% better to come back to it in the discussion if we want to say more.
However, since the effects of rescaling are not fully understood, scaling factors should be used as conservatively as practicable.
The framework provided by \texttt{stdpopsim} should enable a deeper inspection of this issue, which we expect will contribute to
best practices used when applying forward-time simulators such as \texttt{SLiM}.

@@

The \stdpopsim command-line interface, by default, outputs citation information
for the models, genetic maps, and simulation engines used in any particular run.
We hope that this feature will encourage users to appropriately acknowledge the
resources used in published work, and encourage authors
publishing demographic models to contribute to our ongoing community-driven development process.
Together with the \stdpopsim version number and the long-term stable identifiers
for population models and genetic maps,
this citation information will result in well-documented and reproducible
simulation workflows. The individual tree sequence files produced by
\stdpopsim also contain complete provenance information including the command
line arguments, operating system environment and versions of key libraries
used.

@@

As an example of the utility of \stdpopsim, we demonstrate how it can be easily used
to perform a fair comparison of popular demographic inference methods.
Although we present comparison of results from several
methods, our aim at this stage is not to provide an exhaustive
evaluation or ranking of these methods. Our hope is instead to demonstrate how \stdpopsim
will facilitate more detailed future explorations of the strengths and weaknesses of the numerous
inference methods that are available to the population genetics community
(see Discussion).

We start by comparing popular methods for estimating
population size histories of single populations and subsequently
show simple examples of multi-population inference.
To reproducibly evaluate and compare the performance of inference methods, we developed
workflows using \texttt{snakemake} \citep{koster2012snakemake},
available from \url{https://github.com/popsim-consortium/analysis},
that allow efficient computing in multicore or cluster environments.
Our workflow generates $R$ replicates of $C$ chromosomes,
% ACS: "sample" could be confusing because we are both sampling 
% by simulation and sampling from a population.  Does "population samples" (below) help?
producing $n$ population samples in each of a total of $R \times C$ simulations for each demographic model.
After simulation,
the workflow prepares input files for each inference method
by grouping all $n \times R \times C$ simulated chromosomes
into a single file.
Each file is then converted into an input file appropriate for each inference method
(such that all inference methods run on the same simulation replicates).
Each of the inference programs are then run in parallel, and finally,
estimates of population size history from each program are plotted.

% ACS: do we need the "Homo sapiens" at top left in the figure?  It breaks the visual symmetry among YRI, CEU, and CHB, and 
% I think it is unnecessary given the population labels and the description in the legend
% Also: should we include "N(t)" after "Population size" in the y-axis label, to drive home its meaning (used in the text and legend)?
\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/HomSap_OutOfAfricaArchaicAdmixture_5R19.pdf}
\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
size over time ($N(t)$) inferred using 4 different methods: \smcpp, \texttt{stairway plot}, and
\MSMC with $n=2$ and $n=8$ samples. Data were generated by simulating
replicate human genomes under the OutOfAfricaArchaicAdmixture\_5R19 model \citep{ragsdale2019models} and using the
HapMapII\_GRCh37 genetic map~\citep{international2007second}. From top to bottom we show estimates for each
of the three populations in the model (YRI, CEU, and CHB). In shades of blue we show the estimated
$N(t)$ trajectories for each of three replicates.
As a proxy for the ``truth'', in black we show inverse coalescence rates
% perhaps add: "(expressed in numbers of individuals)"    Seems useful to make the units clear
as calculated from the demographic model used for simulation (see text).
}
\label{fig:n_t_ragsdale}
\end{center}
\end{figure}

% ACS: there was indeed a lot of redundancy here.  DRS improved it a lot, but I've tried to do a bit more
\subsubsection*{ Single-population demographic models.}
For single-population demographic models, we compared
\MSMC~\citep{schiffels2014inferring}, \smcpp~\citep{terhorst2017robust}, and
\stairwayplot~\citep{liu2015exploring}
 on simulated genomes sampled from a single population,
under several of the demographic models described above.
However, these experiments raise the question of what to use as the ``true" population sizes in the case of multi-population models with migration.
In particular, when simple single-population models are fit to data simulated under such complex models, 
they are not expected to recover the actual simulated population sizes owing to model misspecification.
Instead, we argue that the best one may expect in such a scenario is to infer a model that accurately reflects the
coalescence time distribution of the simulated model.  In turn, the inverses of these coalescence times (expressed in appropriate units) 
correspond to population sizes.
Under a multi-population model, the coalescence time distribution is influenced by migration between the target population and populations not analyzed in inference,
as well as by the ancestral effective population sizes.
We thus analytically computed inverse coalescence
rates in \texttt{msprime} for each simulated model, and used them as benchmarks for the ``true'' effective population sizes.
See the Appendix for a precise definition and description of the inverse coalescence rate computation.

Figure \ref{fig:n_t_ragsdale} presents the results
from simulations under OutOfAfricaArchaicAdmixture\_5R19,
a model of human migration out of Africa that includes archaic admixture
\citep{ragsdale2019models}, along with an empirical genetic map. In each column of this figure
we show the inferred population size history (denoted $N(t)$)
from samples taken from each of the three extant populations in the model.
In each row we show comparisons among the methods (including two sample sizes for \MSMC).
Blue lines show estimates from each of three replicate whole genome simulations,
and black lines indicate the ``true" values depicted by the inverse coalescence rates (although in this specific model
the inverse coalescence rates are very close to the simulated population sizes;
Figure \ref{fig:census_vs_IRC}).
%DRS: trying to cut down on the redundancy between the coal rate stuff here and the previous paragraph
While there is variation in accuracy among methods, populations, and individual replicates,
the methods generally produce a good estimate
of the true effective population sizes of the simulations, with
inferred values mostly within a factor of two of the truth,
and most methods inferring a bottleneck at approximately the correct time.

@@

We can also
see how well methods might do at recovering the population history of a constant-sized population,
with human genome architecture and genetic map.
We show results of such an
experiment in Figure \ref{fig:n_t_HomSap_constant}.
All methods recover population size within a factor of two of the simulated values, however
SMC-based methods
% ACS: suggest eliminating this vague conjecture.  Which regularization exactly is meant?
%perhaps due to their regularization, 
tend to infer sinusoidal
patterns of population size even though no such change is present.

@@

\subsubsection*{Multi-population demographic models.}
As \stdpopsim implements multi-population demographic models, we also
explored parameter estimation of population divergence parameters. In particular,
we simulated data under multi-population models for humans and \textit{D.~melanogaster}
and then inferred  parameters using \dadi, \fastsimcoal, and \smcpp.
For simplicity, we conducted inference in \dadi and \fastsimcoal by fitting an isolation with migration (IM) model
with constant population sizes and bi-directional migration \citep{hey2004im}.
Our motivation for fitting this simple
IM model was to mimic the typical approach of two population inference on empirical
data, where the user is not aware of the `true' underlying demography and the inference
model is often misspecified.
For human models with more than two populations \citep[e.g.,][]{gutenkunst2009inferring}
this limitation means that users are inferring parameters for a model that does
not match the model from which the data were generated (Figures
\ref{fig:IM_popn_human}A and B). However, since the model used for inference also
allows gene flow between populations, we directly compare estimated effective population sizes
to the values used in simulations (black line in Figure \ref{fig:IM_popn_human}C)
and not the inverse coalescence rates.

In Figure \ref{fig:IM_popn_human}C we show estimates of population sizes and divergence
time, for each of the inference methods, using samples drawn from African and European populations
simulated under the OutOfAfrica\_3G09 model. Our results highlight many
of the strengths and weaknesses of the different methods.
For instance, the SFS-based approaches with simple IM models do not capture
recent exponential growth in the CEU population, but do consistently recover the
simulated YRI population size history. Moreover, these approaches allow 
migration rates to be estimated (Figure \ref{fig:homsap_mig_rates}), and lead to more accurate inferences
of divergence times. However, these migration rate estimates are somewhat biased.
%ACS: seems unnecessary and redundant
%likely due to model misspecification.
In contrast, \smcpp is much better at capturing the recent exponential
growth in the CEU population, though the inferred population sizes are generally noisier.
In addition, \smcpp consistently underestimates divergence times
because it assumes no migration between populations (Figure \ref{fig:IM_popn_human}C).

Again, we can extend this analysis to other taxa and examine the performance of these methods
for a two-population model of \textit{D.~melanogaster}. Figure
\ref{fig:two_popn_fly} shows inference results using data simulated under
the OutOfAfrica\_2L06 model. This model includes an ancestral population in
Africa from which a European population splits off following a bottleneck, with no
post-divergence gene flow between the African and European population (Figure \ref{fig:two_popn_fly}A).
Here again, we find that \dadi and \fastsimcoal infer more consistent histories,
but they do not detect the brief bottleneck in Europe, due to the inference model
not allowing for population size changes after the population split. In addition, \dadi
and \fastsimcoal both do reasonably well at correctly inferring the absence of
migration (Figure \ref{fig:dromel_mig_rates}). In contrast, the inferred
demographic parameters from \smcpp are more noisy, though in some cases better
capture the short bottleneck in the European population.
